%% Set the space above the title appropriately, for the amount of text
%% that you have in your abstract

\vspace*{1.5cm}
%%\begin{center}

{\Huge\textbf{Acknowledgements}}
%%\end{center}

\vspace*{1.5cm}
%%\normalsize
%\begin{abstract}

I would like, first and fore most, to thank all my supervisors, Noah Russell, Tomas Bellamy and Stephen Coombes for giving me the opportunity to engage in this fascinating and challenging project, for allowing me to take part in an international training network where I received a myriad of invaluable experiences and for providing me with all the conditions to succeed.

I would especially like to thank Dr. Federico Dajas-Bailador for his interest and involvement in my project and for providing truly useful insights and advice.

I would like to express my gratitude to my friends and colleagues in the lab, to Raquel Ribeiro, Sebastiaan Van Nuffel and Bo Fu for our close collaborations and for being there in times of need, to Carmel Howe and Jamie Williams for being such charming people and making me feel welcome in this (not so) foreign land and to Alex Johnstone, Benjamin Paul, Chris Towlson, Dee Thakker, Hala Dhowre, Houda Sahaf, Katharina Reusch, Yuezhao Li (Kimi), Solomon Idinyang, Sunil Rajput, Teresa Lobato de Almeida and everyone else for sharing this unique experience and for teaching me so much.

I would also like to thank my friends and colleagues in Maths Agne Tilunaite, Alban Levy, David Hawker and Mayte Bonilla and Sid Visser for interesting and fruitful conversations and for always lending an ear as well as advice.

I feel indebted to our friends here in the UK, it would not have been the same without all those family weekends and homely dinners with Inma, Andreas and Matias, Thao, Huang and Susu, Nela, Nuno and Jamie and Paula, Amael and Lara.

I would like to send deep appreciation to my father and his partner Illana as well as my siblings Sharon and Yahel for always standing by to catch me. Your support and presence have been instrumental in allowing me to overcome the obstacles I faced. At the same time I would like to express my thankfulness to my friends from Israel Uri Cammey, Mark Shein, Noah Levine-Small Ran Sofer, Shay Ariely and Tamir Ezer. Our frequent conversations have been a solid rock in my ever changing life.

Lastly and most importantly, I would to send fervent gratitude and love to my wife Hagit and daughter Amit. Hagit, our relationship has been tested time and again. Your resilience and understanding in helping me to do what was necessary to go through this Ph.D and to write this thesis will stay with us forever. Amit, thank you for being the lovely and astounding creature that you are.
